---
title: "p8105_hw2_xm2356"
author: "Xinyin Miao (xm2356)"
date: "2025-09-27"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(readxl)
library(knitr)
```

# Problem 1

## Clean data in pols-month

```{r}
## load data
pols_df = read_csv("data/fivethirtyeight_datasets/pols-month.csv")

## clean data
pols_cleaned = pols_df |> 
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) |> 
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    day = as.integer(day),

    month = case_when(
      month == 1 ~ "January", month == 2 ~ "February", month == 3 ~ "March",
      month == 4 ~ "April", month == 5 ~ "May", month == 6 ~ "June",
      month == 7 ~ "July", month == 8 ~ "August", month == 9 ~ "September",
      month == 10 ~ "October", month == 11 ~ "November", month == 12 ~ "December",
      TRUE ~ as.character(month)
    ),
    president = case_when(
      prez_dem == 1 ~ "dem",
      prez_gop == 1 ~ "gop",
      TRUE ~ NA_character_
    )
  ) |> 
  select(-prez_dem, -prez_gop, -day) |> 
  select (year, month, president, everything())
  
pols_cleaned
```

The pols-month.csv dataset contained monthly political control data, including presidential party affiliation (prez_gop/prez_dem) and the distribution of governorships and congressional seats, starting from 1947. 

## Clean data in snp

```{r}
## load data
snp_df = read_csv("data/fivethirtyeight_datasets/snp.csv")

## clean data
snp_cleaned = snp_df |> 
  separate(date, into = c("month_num", "day", "year_short"), sep = "/", convert = TRUE)  |> 

  mutate(
    year = case_when(
      year_short <= 49 ~ year_short + 2000,
      year_short >= 50 ~ year_short + 1900,
    ),
    month = case_when(
      month_num == 1 ~ "January", month_num == 2 ~ "February", month_num == 3 ~ "March",
      month_num == 4 ~ "April", month_num == 5 ~ "May", month_num == 6 ~ "June",
      month_num == 7 ~ "July", month_num == 8 ~ "August", month_num == 9 ~ "September",
      month_num == 10 ~ "October", month_num == 11 ~ "November", month_num == 12 ~ "December",
      TRUE ~ NA_character_
      )
  )  |>  
  select(year, month, snp_value = close) |> 
    # arrange by year and month
  arrange(year, month)

snp_cleaned
```

The snp.csv file provided the monthly closing value of the S&P 500 index, tracking market performance across the period. 

## Clean data in unemployment

```{r}
## load data
unemployment_df = read_csv("data/fivethirtyeight_datasets/unemployment.csv")

## tidy data
unemployment_cleaned = unemployment_df |> 
  rename(year = Year) |> 
  pivot_longer(
    cols = Jan:Dec,
    names_to = "month_abbr",
    values_to = "unemployment_rate"
  ) |> 
mutate(
    month = case_when(
      month_abbr == "Jan" ~ "January", month_abbr == "Feb" ~ "February",
      month_abbr == "Mar" ~ "March", month_abbr == "Apr" ~ "April",
      month_abbr == "May" ~ "May", month_abbr == "Jun" ~ "June",
      month_abbr == "Jul" ~ "July", month_abbr == "Aug" ~ "August",
      month_abbr == "Sep" ~ "September", month_abbr == "Oct" ~ "October",
      month_abbr == "Nov" ~ "November", month_abbr == "Dec" ~ "December",
      TRUE ~ NA_character_
    )
) |> 
select(year, month, unemployment_rate) |> 
arrange(year, month)

unemployment_cleaned
```

The unemployment.csv dataset held the U.S. unemployment rate, originally in a wide format with each month as a separate column. After cleaning and merging, the resulting dataset has 816 observations and 3 variables.

## Join the data

```{r}
merged_df = pols_cleaned |> 
  left_join(snp_cleaned, by = c("year", "month")) |> 
  left_join(unemployment_cleaned, by = c("year", "month"))

merged_df
final_dim = dim(merged_df)
final_years = range(merged_df$year)
final_names = names(merged_df)

final_dim
final_years
final_names
```

This project merged three datasets provided by FiveThirtyEight.
The resulting tidy dataset, created by merging the three files using left_join, has 822 observations and 11 variables. The data spans the years from 1947 to 2015. Key variables include year, month, president, snp_value, and unemployment_rate.


# Problem 2


```{r}
trash_file <- "data/202409 Trash Wheel Collection Data.xlsx"

mr_trash_df <- read_excel(
  trash_file,
  sheet = "Mr. Trash Wheel",
  range = cell_cols("A:N")
) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    sports_balls = as.integer(round(sports_balls, 0)),
    year = as.numeric(year),    
    wheel = "Mr. Trash Wheel"
  )

prof_trash_df <- read_excel(
  trash_file,
  sheet = "Professor Trash Wheel",
  range = cell_cols("A:M")
) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    year = as.numeric(year),   
    wheel = "Professor Trash Wheel"
  )

gwynnda_df <- read_excel(
  trash_file,
  sheet = "Gwynnda Trash Wheel",
  range = cell_cols("A:L")
) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(
    year = as.numeric(year),     
    wheel = "Gwynnda Trash Wheel"
  )

trash_wheel_df <- bind_rows(mr_trash_df, prof_trash_df, gwynnda_df)

n_obs <- nrow(trash_wheel_df)


```

## The total weight of trash collected by Professor Trash Wheel
```{r}
prof_total_weight <- trash_wheel_df |> 
  filter(wheel == "Professor Trash Wheel") |> 
  summarise(total = sum(weight_tons, na.rm = TRUE)) |> 
  pull(total)
```

## The total number of cigarette butts collected by Gwynnda in June of 2022
```{r}
gwynnda_cigs_june2022 <- trash_wheel_df |> 
  filter(wheel == "Gwynnda Trash Wheel",
         year == 2022, month(date) == 6) |> 
  summarise(total = sum(cigarette_butts, na.rm = TRUE)) |> 
  pull(total)
```

## Paragraph about these data
```{r}
glue::glue("
The combined Trash Wheel dataset contains {n_obs} observations from three 
sources: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. 
Key variables include date of collection, weight in tons, volume in cubic yards, 
and counts of items such as plastic bottles, cigarette butts, and sports balls. 
For example, Professor Trash Wheel has collected a total of {prof_total_weight} tons of trash, 
while Gwynnda collected {gwynnda_cigs_june2022} cigarette butts in June 2022.
")

```


# Problem 3

```{r}

zori_file <- "data/zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv"
zip_file  <- "data/zillow_data/Zip Codes.csv"


zori_df <- read_csv(zori_file) |> 
  pivot_longer(
    cols = matches("^20"),  
    names_to = "month",
    values_to = "zori"
  ) |> 
  mutate(
    month = ymd(month),      
    year = year(month),
    month_num = month(month)
  )

zip_df <- read_csv(zip_file) |> 
   janitor::clean_names() 


```








